Integer Programming Formulations for the Shared Multicast Tree Problem

### 2018-12-27 smt/ivanova.tex
MI: Remarks from minor revision incorporated - mostly typos and grammar.

\bibitem[Bauer et al.(2008)]{bauer08}
Bauer J, Haugland D, Yuan D (2008)
Analysis and computational study of several integer programming formulations for minimum-energy multicasting in wireless ad hoc networks.
Networks 52(2):57–68
\textcolor{blue}{\bibitem[Bein and Zheng(2010)]{bein10}
Bein D, Zheng SQ (2010)
Energy efficient all-to-all broadcast in all-wireless networks.
Information Sciences 180(10):1781--1792}
\textcolor{blue}{\bibitem[Bhukya and Singh(2014)]{bhukya14}
Bhukya WN, Singh A (2014)
\emph{p}-shrink: A Heuristic for Improving Minimum All-to-All Power Broadcast Trees in Wireless Networks.
In: Proceedings of Ninth International Conference on Wireless Communication and Sensor Networks 2014 
Vol.\ 299, pp 61--69}

@@

\bibitem[Guo and Yang(2004)]{guo}
\textcolor{blue}{Guo S, Yang O (2004)
Minimum Energy Multicast Routing for Wireless Ad-hoc Networks with Adaptive Antennas 
In: Proceedings of the 12th IEEE International Conference on Network Protocols ICNP 2004,
pp 151--160}

@@

\bibitem[Montemanni and Leggieri(2011)]{montemanni11}
Montemanni R, Leggieri V (2011)
A branch and price algorithm for the minimum power multicasting problem in wireless sensor networks.
Math Meth Oper Res
74:327--342
\textcolor{blue}{\bibitem[Pajor et al.(2018)]{pajor18}
Pajor T, Uchoa E, Werneck RF (2018)
A robust and scalable algorithm for the Steiner problem in graphs.
Math Prog Comp 10:69--118
}

>>>>

\bibitem[Bauer et al.(2008)]{bauer08}
Bauer J, Haugland D, Yuan D (2008)
Analysis and computational study of several integer programming formulations for minimum-energy multicasting in wireless ad hoc networks.
Networks 52(2):57–68

\bibitem[Bein and Zheng(2010)]{bein10}
Bein D, Zheng SQ (2010)
Energy efficient all-to-all broadcast in all-wireless networks.
Information Sciences 180(10):1781--1792

\bibitem[Bhukya and Singh(2014)]{bhukya14}
Bhukya WN, Singh A (2014)
\emph{p}-shrink: A Heuristic for Improving Minimum All-to-All Power Broadcast Trees in Wireless Networks.
In: Proceedings of Ninth International Conference on Wireless Communication and Sensor Networks 2014 
Vol.\ 299, pp 61--69

@@

\bibitem[Guo and Yang(2004)]{guo}
Guo S, Yang O (2004)
Minimum Energy Multicast Routing for Wireless Ad-hoc Networks with Adaptive Antennas 
In: Proceedings of the 12th IEEE International Conference on Network Protocols ICNP 2004,
pp 151--160

@@

\bibitem[Montemanni and Leggieri(2011)]{montemanni11}
Montemanni R, Leggieri V (2011)
A branch and price algorithm for the minimum power multicasting problem in wireless sensor networks.
Math Meth Oper Res
74:327--342

\bibitem[Pajor et al.(2018)]{pajor18}
Pajor T, Uchoa E, Werneck RF (2018)
A robust and scalable algorithm for the Steiner problem in graphs.
Math Prog Comp 10:69--118

### 2018-12-27 smt/sec1-intro.tex
MI: Remarks from minor revision incorporated - mostly typos and grammar.

Wireless ad hoc networks (WANETs) are suitable in various practical applications in both civil and military sector, where a communication infrastructure is either absent or cannot be relied on.
Owing to their quick deployment and simple configuration, WANETs are useful in emergency operations such as natural disaster relief and military command and control.
Recently, the WANET concept has been introduced in ad hoc smart lighting in homes and in offices as well as ad hoc street light networks, where the control includes adjusting dimmable lights.

@@

\textcolor{blue}{\subsection{Related problems}}

A well-known and extensively studied problem in the network design literature is the \textsc{Minimum Energy Broadcast} (MEB) problem.
Given a set of wireless devices with one designated source node, the goal is to assign powers to individual nodes.
The power assignments determine the communication ranges, and induce a broadcast tree such that a signal initiated by the source reaches, directly or via other nodes,
all the remaining nodes.
In the MEB problem, the task is to find power assignments enabling such a signal broadcast while minimizing the total energy consumption.
A natural and frequently addressed extension of MEB, is the \textsc{Minimum Energy Multicast} (MEM) problem.
Analogous to the difference between the \textsc{Minimum Spanning Tree} and the \textsc{Minimum Steiner Tree} problems,
MEM is distinguished from MEB in that not all nodes in the network are destinations of the multicast message.
Hence, the more general problem amounts to finding a Steiner tree spanning the destination nodes such that the sum of the power assignments is minimized.
Non-destination nodes are allowed in the Steiner tree, and will be included to the extent that they contribute to reduced power consumption.

In certain practical settings, MEB fails to reflect reality adequately.
Consider an application where more than one node can act as a source.
% Every node may initiate a message intended for the remaining nodes.
The optimal broadcast trees corresponding to two different sources are not necessarily identical,
and the optimal broadcast trees must be calculated separately for every possible source node.
In order to relay a received signal correctly, the nodes must be able to recognize which node initiated the received signal, and therefore also determine which broadcast tree to use.
Further, which power level to be set must be assessed by the relaying node.
Such overhead calculations require additional energy, and enabling the communication devices to accomplish them may even be impractical.

@@

\textcolor{blue}{\subsection{Contributions}}

Contributions from the present work include two sets of IP models for SMT, presented in order of increasing lower bounding capabilities.
By rigorous analysis of the models, we prove relations between the bounds offered by the corresponding continuous relaxations.
Intended for instances beyond the size that can be solved to optimality, we also develop a constraint generation procedure by which strong bounds can be computed.

For the same reasons that MEB has been generalized to MEM, it is natural and desirable to extend SBT to its multicast equivalent.
The \textsc{Minimum Shared Multicast Tree} (SMT) problem is thus at the forefront of this paper,
and neither MEB, MEM, nor RAP will be considered further.
Focus is henceforth on instances in which some of the nodes neither initiate new messages nor need messages from others.
Utilizing such \emph{non-destination} nodes to forward messages can in many cases contribute to substantial reductions in the total power consumption.
Non-destination nodes thus have the potential to play the role of Steiner nodes in the multicast tree.
All devices that can initiate a transmission, referred to as \emph{destination} nodes, also have to receive every message.

\textcolor{blue}{\subsection{Structure of the paper}}

The remainder of this paper is organized as follows: 
\textcolor{blue}{Section \ref{sec:literature} summarizes relevant literature and previous works.}
%Section \ref{sec:SBT} describes the SMT problem and gives detailed explanation of its objective function.
\textcolor{blue}{Section \ref{sec:SBT} divided into three subsections introduces the notation, explains the network model, and gives formal definition of the problem.}
Integer linear programming formulations, valid inequalities and their analysis are presented in Section \ref{sec:ILP},
followed by Section \ref{sec:comp}, where the models under study are compared with respect to strength.
In Section \ref{sec:strong}, the models are further extended by variables in higher dimensional spaces,
and a constraint generation procedure is developed.
Computational results are reported in Section \ref{sec:exp},
and suggestions to future work are summarized in Section \ref{sec:conclusion}.

>>>>

Wireless ad hoc networks (WANETs) are suitable in various practical applications in both civil and military sector, 
where a communication infrastructure is either absent or cannot be relied on.
Owing to their quick deployment and simple configuration, WANETs are useful in emergency operations such as natural disaster relief and military command and control.
Recently, the WANET concept has been introduced in ad hoc smart lighting in homes and in offices, 
as well as ad hoc street light networks, where the control includes adjusting dimmable lights.

@@

\subsection{Related problems}

A well-known and extensively studied problem in the network design literature is the \textsc{Minimum Energy Broadcast} (MEB) problem.
Given a set of wireless devices with one designated source node, the goal is to assign power to individual nodes.
The power assignments determine the communication ranges, and induce a broadcast tree such that a signal initiated by the source reaches, directly or via other nodes,
all the remaining nodes.
In the MEB problem, the task is to find power assignments enabling such a signal broadcast while minimizing the total energy consumption.
A natural and frequently addressed extension of MEB, is the \textsc{Minimum Energy Multicast} (MEM) problem.
Analogous to the difference between the \textsc{Minimum Spanning Tree} and the \textsc{Minimum Steiner Tree} problems,
MEM is distinguished from MEB in that not all nodes in the network are destinations of the multicast message.
Hence, the more general problem amounts to finding a Steiner tree spanning the destination nodes such that the sum of the power assignments is minimized.
Non-destination nodes are allowed in the Steiner tree, and will be included to the extent that they contribute to reduced power consumption.

In certain practical settings, MEB fails to reflect reality adequately.
Consider an application where more than one node can act as a source.
% Every node may initiate a message intended for the remaining nodes.
The optimal broadcast trees corresponding to two different sources are not necessarily identical,
and the optimal broadcast trees must be calculated separately for every possible source node.
In order to relay a received signal correctly, the nodes must be able to recognize which node initiated the received signal, and therefore also determine which broadcast tree to use.
Further, which power level to be set must be assessed by the relaying node.
Such overhead calculations require additional energy, and enforcing the communication devices to accomplish them may even be impractical.

@@

\subsection{Contributions}

Contributions from the present work include two sets of Integer Programming (IP) models for SMT, presented in order of increasing lower bounding capabilities.
By rigorous analysis of the models, we prove relations between the bounds offered by the corresponding continuous relaxations.
Intended for instances beyond the size that can be solved to optimality, we also develop a constraint generation procedure by which strong bounds can be computed.

When proven optimality is out of reach, effective methods for computing near-optimal solutions are crucial.
A heuristic method tailored for SMT is another contribution from the current text,
which thus demonstrates how recent progress on the \textsc{Minimum Steiner Tree} problem can be transferred to the problem under study.

\subsection{Scope and structure of the paper}

For the same reasons that MEB has been generalized to MEM, it is natural and desirable to extend SBT to its multicast equivalent.
The \textsc{Minimum Shared Multicast Tree} (SMT) problem is thus at the forefront of this paper,
and neither MEB, MEM, nor RAP will be given much attention.
Focus is henceforth on instances in which some of the nodes neither initiate new messages nor need to receive messages from others.
Utilizing such \emph{non-destination} nodes to forward messages can in many cases contribute to substantial reductions in the total power consumption.
Non-destination nodes thus have the potential to play the role of Steiner nodes in the multicast tree.
All devices that can initiate a transmission, referred to as \emph{destination} nodes, also have to receive every message.

The remainder of this paper is organized as follows: 
Section \ref{sec:literature} summarizes relevant literature and previous works.
Section \ref{sec:SBT} introduces the notation, explains the network model, and gives a formal definition of the problem.
Integer linear programming formulations, valid inequalities and their analysis are presented in Section \ref{sec:ILP},
followed by Section \ref{sec:comp}, where the models under study are compared with respect to their strength.
In Section \ref{sec:strong}, the models are further extended by variables in higher dimensional spaces,
and a constraint generation procedure is developed.
A fast method for computing good, but not necessarily optimal solutions is given in Section \ref{sec:heur}.
Computational results are reported in Section \ref{sec:exp},
and suggestions to future work are summarized in Section \ref{sec:conclusion}.
\newline

### 2018-12-27 smt/sec2-literature.tex
MI: Remarks from minor revision incorporated - mostly typos and grammar.

The combinatorial optimization literature is already rich on Integer Programming (IP) approaches to power minimization in wireless networks.
Since the work of \citet{Wieseltier00onthe}, in particular the MEB and MEM problems have been studied extensively.
Noteworthy contributions to this research include the IP models studied by \citet{das03},
\citet{altinkemer05}, \citet{yuan05}, \citet{yuan08}, \citet{bauer08}, \citet{montemanni11} and \citet{leggieri08}. 
\textcolor{blue}{
The latter paper proposes an IP formulation of MEM based on a set covering model together with a preprocessing procedure reducing the number of constraints.}
\citet{cagalj02} prove that MEB, and thereby MEM, is NP-hard.
Numerous contributions to fast, but possible inexact methods for MEM have thus appeared, and the interested reader is referred to \citep{hsiao13} for an overview. 
\textcolor{blue}{Another study concerning MEM is the work by \cite{guo}, who investigate energy conservation while using adaptive antennas. 
Unlike omnidirectional antennas, adaptive antennas can concentrate their transmission power in a smaller angle.}

\textcolor{blue}{\citet{bein10} study energy efficiency in wireless networks where all nodes can initiate a message, and the communication takes place in a single all-to-all broadcast tree.
A heuristic algorithm generating all-to-all broadcast trees is proposed by \citet{bhukya14}, along with an experimental evaluation confirming its merit.}

>>>>

The combinatorial optimization literature is already rich on IP approaches to power minimization in wireless networks.
Since the work of \citet{Wieseltier00onthe}, in particular the MEB and MEM problems have been studied extensively.
Noteworthy contributions to this research include the IP models studied by \citet{das03},
\citet{altinkemer05}, \citet{yuan05}, \citet{yuan08}, \citet{bauer08}, \citet{montemanni11} and \citet{leggieri08}. 
The latter paper proposes an IP formulation of MEM based on a set covering model, together with a preprocessing procedure reducing the number of constraints.
\citet{cagalj02} prove that MEB, and thereby MEM, is NP-hard.
Numerous contributions to fast, but possibly inexact methods for MEM have thus appeared, and the interested reader is referred to \citep{hsiao13} for an overview. 
Another study concerning MEM is the work by \cite{guo}, who investigate energy conservation while using adaptive antennas. 
Unlike omnidirectional antennas, adaptive antennas can concentrate their transmission power in a smaller angle.

\citet{bein10} study energy efficiency in wireless networks where all nodes can initiate a message, and the communication takes place in a single all-to-all broadcast tree.
A heuristic algorithm generating all-to-all broadcast trees is proposed by \citet{bhukya14}, along with an experimental evaluation confirming its merit.

### 2018-12-27 smt/sec3-sbtsmt.tex
MI (SMT): Minor changes in several files, mainly in Sect 8 - description of the results.
MI: some typos fixed
MI: Remarks from minor revision incorporated - mostly typos and grammar.

\section{The Minimum Shared Multicast Tree Problem}
\label{sec:SBT}

@@

Sending a message directly from (to) node $i$ to (from) node $j$ requires a certain \emph{transmission power} $p_{ij}=p_{ji}$.
Whenever appropriate, $p_{ij}$ is referred to as the \emph{cost} of $\{i,j\}$,
and edges with higher cost than others are said to be \emph{more expensive}.
When the nodes are embedded in the plane, $p_{ij}$ is often assumed \citep{Papadimitriou06SBT,Haugland12Dual,halgamuge} to be
proportional to the Euclidean distance between $i$ and $j$ raised to a power between 2 and 4.
For simplicity, we assume for all $i\in V$ that $p_{ij}\neq p_{ik}$ when $j\neq k$, and define
$W_{ij}=\left\{k\in V_i: p_{ik}\geq p_{ij}\right\}$
as the set of nodes requiring no less power at $i$ than $j$ does.
The notation $G'\subseteq G$ means that $G'=(V_{G'},E_{G'})$ is a subgraph of $G$, i.e., $E_{G'}\subseteq E$ consists of the edges in $E$ with both end nodes in $V_{G'}\subseteq V$.

Let $y \in \{0,1\}^E$ and $X\in\{0,1\}^{A\times D}$ be binary vectors with components corresponding to
edges and pairs of arcs and destination nodes, respectively.
The vector $X^s$ then consists of all components of $X$ corresponding to a pair of which $s$ is the destination.
The subgraph of $G$ \emph{induced by} $y$ is defined as $G_y=(V,E_y)$, where $E_y=\left\{\{i,j\}\in E: y_{ij}=1\right\}$.
Directed graphs induced by binary vectors over $A$ are defined analogously.

@@

\textcolor{blue}{\subsection{Network model} \label{sec:netmod}}

To illustrate the problem under study, consider the tree depicted in Fig. \ref{fig:objexp}.
Six destination nodes ($D=\{i_1,i_2,a,b,c,d\}$) are spanned by use of three Steiner nodes ($V\setminus D = \{e,i,i_3\}$).
Consider node $i$ and its three neighbour nodes $i_1$, $i_2$ and $i_3$, ordered by decreasing power requirement at $i$.
That is, $p_{ii_1}>p_{ii_2}>p_{ii_3}$.
If the transmitting node is outside the shaded area ($i_2$, $c$, or $d$), then node $i$ relays the signal to two of its neighbors, \emph{including $i_1$}.
Exploiting the wireless advantage, node $i$ reaches both neighbors if it is assigned power level $p_{ii_1}$.
Otherwise, if the signal source is either $i_1$, $a$, or $b$, node $i$ passes the signal on to neighbors $i_2$ and $i_3$.
This is accomplished by power $p_{ii_2}$.

\begin{figure}[h!]
\centering
\begin{tikzpicture}[scale=.6]
 \draw[rounded corners=5mm, fill=gray!20, draw=white] (-.7,-.2) -- (3.5,0.7) -- (0.2,4.2) -- cycle;
 \begin{scope}[every node/.style={circle,draw=black,fill=gray!60,minimum size=1.5em, inner sep=2, scale=.8}]
  \node (a) at (0,0.5) {$a$};
  \node (b) at (0.5,3) {$b$};
  \node (c) at (9,2.5) {$c$};
  \node (d) at (9.5,0.9) {$d$};
  \node[scale=.85] (i1) at (2,1) {$i_1$};
  \node[scale=.85] (i2) at (4,3.1) {$i_2$};
  \node[scale=.85,fill=none] (i3) at (5.8,1.2) {$i_3$};
  \node[fill=none] (i) at (4.5,1.5) {$i$};
  \node[fill=none] (e) at (8,1.5) {$e$};
 \end{scope}
 \draw (a) -- (i1) -- (b);
 \draw (i1) -- (i) -- (i3) -- (e) -- (c);
 \draw (e) -- (d);
 \draw (i) -- (i2);
\end{tikzpicture}
\caption{A multicast tree illustrating the cost contribution from node $i$ to the objective function.
	 Destinations and Steiner nodes are denoted by filled and empty circles, respectively.}
\label{fig:objexp}
\end{figure}

@@

\textcolor{blue}{
The cost model outlined above aims for minimization of the transmission energy.
As pointed out by \citet{halgamuge}, transmission constitutes a substantial part of the total energy consumption in wireless sensor networks.
The energy consumed by other operations, such as sensing, logging, processing, actuation, and cluster formation is less sensitive to the topology
of the multicast tree, and is neglected in the current work.
This is in line with the approach taken in the wide range of literature on optimization models for wireless networks mentioned in Section \ref{sec:literature}.
For a detailed energy model for wireless sensor networks, the reader is referred to the article by \citet{halgamuge}.}

@@

Problem \ref{def:problem} also has close resemblance with the SBT problem, which is the special case of SMT where $D=V$.
Like many other network design problems, \textsc{Minimum Shared Multicast Tree} is NP-hard.
This follows directly from the NP-hardness of SBT \citep{Papadimitriou06SBT}.

>>>>

\section{The Minimum Shared Multicast Tree Problem}
\label{sec:SBT}
Like in studies of other problems in combinatorial optimization,
a theoretical investigation of SMT requires abstraction and simplification of the real-world communication network.

@@

Sending a message directly from (to) node $i$ to (from) node $j$ requires a positive \emph{transmission power} $p_{ij}=p_{ji}$.
Whenever appropriate, $p_{ij}$ is referred to as the \emph{cost} of $\{i,j\}$,
and edges with higher cost than others are said to be \emph{more expensive}.
When the nodes are embedded in the plane, $p_{ij}$ is often assumed \citep{Papadimitriou06SBT,Haugland12Dual,halgamuge} to be
proportional to the Euclidean distance between $i$ and $j$ raised to a power between 2 and 4.
For simplicity, we assume for all $i\in V$ that $p_{ij}\neq p_{ik}$ when $j\neq k$, and define
$W_{ij}=\left\{k\in V_i: p_{ik}\geq p_{ij}\right\}$
as the set of nodes requiring no less power at $i$ than $j$ does.
The notation $G'\subseteq G$ means that $G'=(V_{G'},E_{G'})$ is a subgraph of $G$, 
i.e., $E_{G'}\subseteq E$ consists uniquely of edges in $E$ with both end nodes in $V_{G'}\subseteq V$.

Let $y \in \{0,1\}^E$ and $X\in\{0,1\}^{A\times D}$ be binary vectors with components corresponding to edges,
and pairs of arcs and destination nodes, respectively.
The vector $X^s$ then consists of all components of $X$ corresponding to a pair of which $s$ is the destination.
The subgraph of $G$ \emph{induced by} $y$ is defined as $G_y=(V,E_y)$, where $E_y=\left\{\{i,j\}\in E: y_{ij}=1\right\}$.
Directed graphs induced by binary vectors over $A$ are defined analogously.

@@

\subsection{Network model} \label{sec:netmod}

To illustrate the problem under study, consider the tree depicted in Fig. \ref{fig:objexp}.
Seven destination nodes ($D=\{i_1,i_2,a,b,c,d,e\}$) are spanned by use of two Steiner nodes ($V\setminus D = \{i,i_3\}$).
Consider node $i$ and its three neighbour nodes $i_1$, $i_2$ and $i_3$, ordered by decreasing power requirement at $i$.
That is, $p_{ii_1}>p_{ii_2}>p_{ii_3}$.
If the transmitting node is outside the shaded area ($i_2$, $c$, $d$, or $e$), then node $i$ relays the signal to two of its neighbors, \emph{including $i_1$}.
Exploiting the wireless advantage, the signal from node $i$ reaches both neighbors if it is assigned power level $p_{ii_1}$.
Otherwise, if the signal source is either $i_1$, $a$, or $b$, node $i$ passes the signal on to neighbors $i_2$ and $i_3$.
This is accomplished by setting the power level of $i$ to $p_{ii_2}$.

\begin{figure}[h!]
\centering
\begin{tikzpicture}[scale=.6]
 \draw[rounded corners=5mm, fill=gray!20, draw=white] (-.7,-.2) -- (3.5,0.7) -- (0.2,4.2) -- cycle;
 \begin{scope}[every node/.style={circle,draw=black,fill=gray!60,minimum size=1.5em, inner sep=2, scale=.8}]
  \node (a) at (0,0.5) {$a$};
  \node (b) at (0.5,3) {$b$};
  \node (c) at (9,2.5) {$c$};
  \node (d) at (9.5,0.9) {$d$};
  \node[scale=.85] (i1) at (2,1) {$i_1$};
  \node[scale=.85] (i2) at (4,3.1) {$i_2$};
  \node[scale=.85,fill=none] (i3) at (5.8,1.2) {$i_3$};
  \node[fill=none] (i) at (4.5,1.5) {$i$};
  \node (e) at (8,1.5) {$e$};
 \end{scope}
 \draw (a) -- (i1) -- (b);
 \draw (i1) -- (i) -- (i3) -- (e) -- (c);
 \draw (e) -- (d);
 \draw (i) -- (i2);
\end{tikzpicture}
\caption{A multicast tree illustrating the cost contribution from node $i$ to the objective function.
	 Destinations and Steiner nodes are denoted by filled and empty circles, respectively.}
\label{fig:objexp}
\end{figure}

@@

The cost model outlined above aims for minimization of the transmission energy.
As pointed out by \citet{halgamuge}, transmission constitutes a substantial part of the total energy consumption in wireless sensor networks.
The energy consumed by other operations, such as sensing, logging, processing, actuation, and cluster formation is less sensitive to the topology
of the multicast tree, and is neglected in the current work.
This is in line with the approach taken in the wide range of literature on optimization models for wireless networks mentioned in Section \ref{sec:literature}.
For a detailed energy model for wireless sensor networks, the reader is referred to the article by \citet{halgamuge}.

@@

Problem \ref{def:problem} also has close resemblance with the SBT problem, which is the special case of SMT where $D=V$.
Like many other network design problems, SMT is NP-hard.
This follows directly from the NP-hardness of SBT \citep{Papadimitriou06SBT}.
